Seismic Time Series Analysis\\Statistics for Spatio-Temporal Data

### 2020-06-12 draft.tex
Last fixes before delivery

The correct estimation of a background noise, correlated or not with human activity, could help institutions to better isolate less prominent earthquakes and similar events from the seismic data. This could result in better understanding of the frequency of earthquakes and by studying their more-precise evolution over time.

\subsubsection{Data source and description}
The Royal Observatory of Belgium (ROB) gathers and publishes \cite{RoyalDataPolicy,Data20200402} data about vertical ground displacement from 9 stations through the country, linking them to seismic events whenever possible \cite{DataEvents20200402}. They also offer a data visualization tool for the same data \cite{DataVisualization20200402}.

@@

Out of the 9 stations, the following 6 have interesting positions, related to possible correlations with human activity:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item \textbf{Uccle}, \textbf{Sart-Tilman}, and \textbf{Ében-Émael} are near a city (Brussels, Liege, and Maastricht).
	\item \textbf{Ostenda} This station is near the city of Ostenda, that borders with the sea.
	\item \textbf{Membach} and \textbf{Dourbes} are near a natural park.
\end{itemize}

@@

\subsection{Data cleanup and transformation}\label{sec:cleanup-transform}
\subsubsection{Data gathering and cleanup}
The data is provided in CSV format, one file per day, with only two columns related to the min-max pair \cite{Data20200402} and no column related to time. This causes two problems:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item For merging days, we need to add a column with the second-precise datetime
	\item In days with missing values there are placeholder values instead a missing row, that would make the following rows misinterpreted for measurements on the wrong second. The placeholder values are one of these two min-max pairs: $(0,0),\, (-1,1)$.
\end{enumerate}

@@

\subsubsection{Data aggregation and initial transformation}
Since we are not interested in a second-precise analysis, and it would be computationally expensive to deal with the whole dataset, we aggregate data by hour. The aggregation results in around 14 thousand points, that can still be captured by interpretable models.

While performing this operation, we should consider the correct time-zone: any seasonal analysis dealing with UTC dates would be out of phase for the half-year of legal time, with respect to the half-year of solar time. With the help of the \texttt{lubridate} R package we converted times to local time-zones, and then we kept them as absolute values, because all the R functions we use work on the UTC conversion of datetimes by default. This transformation made it possible to have each hour in the timeseries to represent the local time in Belgium (\textit{Europe/Brussels}), without being influenced by the legal time.

The aggregation was performed considering the mean total ground displacement, for the following reasons:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item When an hour contains some missing seconds, these do not heavily affect the aggregation result, in contrast with summing the values
	\item Seismic events, such as quakes and mine explosions, normally last some minutes at most, and will affect the result only slightly, as opposed to the hourly maximum
	\item Usually, maximum value aggregation does not follow the Normal distribution, while some models that we are going to use assume the data follow a Normal distribution
\end{itemize}

@@

Since our main focus is on the seasonalities themselves, we aim to prevent assumptions based on this objective. In other words, we want to avoid searching for correlation with human activity while imposing human-based seasonality periods. Thus, we use an analytical method to extract seasonalities that the data itself suggests to us. Still, we recognize that the already performed transformations impose an human-based time alteration to the data itself (Section \ref{sec:cleanup-transform}).

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/whole data plot-2"}
		\caption{Uccle station}
		\label{fig:zoom:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/whole data plot-2"}
		\caption{Membach station}
		\label{fig:zoom:mems}
	\end{subfigure}
	\caption{Zoom from January $6^{th}$ to February $3^{rd}$ 2019}
	\label{fig:zoom}
\end{figure}
%
The method we are going to apply is based on the Fourier Transform (FT), and generates a \textit{periodgram} highlighting the frequencies that might correspond to a seasonality component \cite[chapter 11]{brockwell1991time}.
Applying the FT for any frequency, we transform our 1-dimensional data into a 2-dimensional data mapped around a circle. Then, we can sum all the resulting vectors (that are our new data points) to obtain a vector for each frequency, whose absolute value represents the power level of the same frequency. An high power level represents a significant periodicity on that frequency in our original data.

A common operation to increase the difference between significant and non-significant periodicities is to apply a Moving Average (MA) smoothing filter with a smaller window than the interesting ones. We will use such a filter with $p=2$.

On this note we must specify that, when a significant periodicity is found, it influences smaller periodicities given by $f*\{2,3,\dots\}$ the significant frequency \cite{brockwell1991time}. Also, if any periodicity is found that is longer than the original dataset, it must be considered as simple chance, rather than serendipity.

@@

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/periodgram-1"}
		\caption{Uccle station}
		\label{fig:periodgram:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/periodgram-1"}
		\caption{Membach station}
		\label{fig:periodgram:mems}
	\end{subfigure}
	\caption{Periodgram results}
	\label{fig:periodgram}
\end{figure}
%
Both highlight multiple 24-hour periodicities. This is given by approximation factors, since we round the divisions result to the nearest natural number. It also highlights that using a single 24-hour seasonality could leave some information on the remaining components. Still, in both stations it is the most significant periodicity.

@@

\subsubsection{Dataset division}
As best practices suggest, we are going to divide the dataset into training and testing datasets with a ratio around $85\%/15\% - 75\%/25\%$.
Since one of our objectives is to assess any significant difference from before and after the Belgium lockdown (on March $14^{th}$ the \textit{soft} lockdown, while on March $18^{th}$ the complete lockdown), we selected data from the start of our timeseries until 20-01-2020 (64 weeks $\approx 82\%$) as training, while the test follows and ends on 27-04-2020 (14 weeks $\approx 18\%$).

@@

This initial division demostrated to be computationally heavy while fitting some models with weekly seasonality, so that it would be challenging to fit more than a couple of models.
Given this issue, we choose the training/testing sets to hold 52/18 weeks: this is the greater train-to-test ratio we deem usable ($\approx 75\%/25\%$), while maintaining computational feasibility.
The assessment of the lockdown influence will be discussed in Section~\ref{sec:methodology}.

Some R commands cannot handle timeseries with missing values, but as we have seen both stations datasets include missing values.
Also, please note that we are going to have a timeseries object indexed by weeks, so that it contains the 168 hours between each integer index.
Since we know that the longest missing values series is 90 hours long, we can apply a simple completion strategy for serving those commands a filled timeseries: we will substitute missing values with the mean of the week they are in; in R, \texttt{na.aggregate(x, floor)}.

\subsubsection{Decomposition}
In order to decompose our dataset into trend and seasonality(-ties) components, we followed these steps:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply a smoothing filter to extract the trend component
	\item Remove the trend from the dataset and extract the seasonal figure using the local mean method
	\item Remove the trend and the seasonal components to obtain the residuals
\end{enumerate}
While doing so, we tested different smoothing filters: simple filter (p=seasonality), Spencer's 15-point filter and MA smoothing filter (f=seasonality). Since they do not produce particularly different seasonalities, we choose to keep the results obtained by applying the MA smoothing filter, for its significance in terms of seasonally adjusted data.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/extract both 24 and 168-1.pdf}
		\caption{Uccle station}
		\label{fig:decomposed:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.mems_files/figure-latex/extract both 24 and 168-1.pdf}
		\caption{Membach station}
		\label{fig:decomposed:mems}
	\end{subfigure}
	\caption{Timeseries decomposition}{\centering\small Only the second method results are shown for clarity. Week 10 includes January $1^{st}$.\\}
	\label{fig:decomposed}
\end{figure}
%
Since the periodgrams suggested the presence of two significative periodicities, we also tested the extraction of both corresponding seasonalities:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply the MA smoothing filter (f=24) on the dataset to extract the trend component
	\item Remove the trend from the dataset and extract the 24-hour seasonal figure using the local mean method
	\item Remove the 24-hour seasonal components to obtain a deseasoned dataset
	\item Apply the MA smoothing filter (f=168) on the deseasoned dataset to extract the updated trend component
	\item Remove the trend from the already deseasoned dataset and extract the 168-hour seasonal figure using the local mean method
	\item Repeat steps 1-5 using the last deseasoned dataset until convergence (2-3 cycles)
	\item Removing the last fitted trend and the two seasonalities we obtain the final residuals
\end{enumerate}

@@

From tha same decomposition results, we notice that data around January $1^{st}$ show a significant drop in values. That is highlighted by both trends, where the overall minimum falls in that period, and by the residuals for Uccle data, showing a clear overestimation.

\section{Methodology}\label{sec:methodology}
Having two different datasets, we will operate on them separately and then we will compare the model interpretation.

@@

The model we fit is the ARIMA model, including the seasonality components. To identify the order of the different components described by ARIMA, we will analyse the autocorrelation and partial autocorrelation of our datasets.
As a reminder:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item non-seasonal AR process of order $p$ is described by an exponentially or sigmoidally decaying ACF, and by a cut-off of significant PACF values at lag $p$
	\item non-seasonal MA process of order $q$ is described by an exponentially or sigmoidally decaying PACF, and by a cut-off of significant ACF values at lag $q$
	\item seasonal AR process of order $P$ is described by an exponential or sigmoidal decay at seasonal lags on ACF values, and by a cut-off of significant PACF values at each seasonal lag until the $P^{th}$
	\item seasonal MA process of order $Q$ is described by an exponential or sigmoidal decay at seasonal lags on PACF values, and by a cut-off of significant ACF values at each seasonal lag until the $Q^{th}$
	\item given how significance is assessed for the ACF/PACF spikes, one over 20 spikes (5\%) could be \textit{coincidentally significant} or \textit{coincidentally \textbf{non}-significant}
\end{itemize}

The same model will be then used to forecast other 8 weeks from March $2^{nd}$, so that we can assess the forecast power over the lockdown period. Given the fact that both our series are quite stable over time (\figurename{s} \ref{fig:decomposed:uccs}, \ref{fig:mems:box-cox:data}), we can expect that the predicted variability for further forecasts will not significantly increase.

@@

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-35}
		\caption{ACF and PACF plots}
		\label{fig:arima-26}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-32}
		\caption{Q-Q plot}
		\label{fig:arima-24}
	\end{subfigure}
	\caption{Plots for $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
	\label{fig:arima}
\end{figure}
%
The ACF and PACF of the residuals for all three models are very similar (\figurename~\ref{fig:arima-26}), even if the models themselves have great differences in meaning and in number of parameters: they highlight a clear seasonal ARMA process with periodicity 24. This shows how our previous conclusion that the 168 hours seasonality could include the 24 hours one is flawed. Unfortunately, the R function we are using does not permit to estimate multiple seasonalities\footnote{\textit{Arima} method from the \textit{forecast} R package, only specifying the seasonal and non-seasonal orders.}.

@@

Also, it must be noted that a January $1^{st}$ period lies in our test dataset, that is already noticed as low-valued period (\figurename~\ref{fig:decomposed}). We can notice a clear overestimation during those couple of weeks (around week 62 on \figurename~\ref{fig:arima-forecast}).

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{project.uccs_files/figure-latex/arima-37}
		\caption{$\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
		\label{fig:arima-forecast:200011}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{project.uccs_files/figure-latex/arima-69}
		\caption{$\mathrm{ARIMA}(3,1,1)(0,1,1)_{168}$}
		\label{fig:arima-forecast:412011}
	\end{subfigure}
	\caption{ARIMA forecasts}{\footnotesize$\mathrm{ARIMA}(4,1,2)(0,1,1)_{168}$ behaves like $\mathrm{ARIMA}(3,1,1)(0,1,1)_{168}$ with faster increasing variablity}
	\label{fig:arima-forecast}
\end{figure}
%
We can see in \figurename~\ref{fig:arima-forecast} the forecast behaviour for our models. We can appreciate the constant span of the confidence intervals in \figurename~\ref{fig:arima-forecast:200011}, in constrast with the more complex models that have increasing intervals further away from the training set. It must be noted that the increasing intervals include negative values, that are actually impossible to obtain given the data meaning (difference between maximum and minimum measurements).

@@

\subsection{Membach station}
\subsubsection{ARIMA}
Similarly to data from Uccle, data from Membach doesn't respect the Normality assumption. We tested if the Q-Q plot showed significantly better approximation of the Normal distribution after applying the \textit{log} or \textit{Box-Cox} transformations, and the latter makes the data approach Normality with $\lambda\approx-1$ (\figurename~\ref{fig:mems:box-cox}). This solves the high peaks problem we already noticed while decomposing to highlight the trend and seasonalities.

@@

In this case, the employment of \textit{auto.arima} stepwise methodology$^{\ref{footnote:auto.arima}}$ highlights the first model improved with one non-seasonal MA parameter more. This is confirmed by the ACF/PACF plots of its residuals. Their analysis for the second model do not lead to any improvement.
The ACF and PACF plots of both models are similar to the ones from the chosen model for the Uccle station (\figurename~\ref{fig:arima-24}), and we can notice the same 24 hours remaining periodicity.

@@

Analytically, we have quite a difference between the two models in terms of adherence of the confidence intervals:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item $\mathrm{ARIMA}(0,1,1)(0,1,1)_{168}$: $\approx98\%$ of data is in the $80\%$ CI and $\approx99\%$ in the $95\%$ CI
	\item $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$: $\approx73\%$ of data is in the $80\%$ CI and $\approx89\%$ in the $95\%$ CI
\end{itemize}
As we see, the first model over-generalises the variability, while the latter seems more precise, even if losing some information. Also considering the MAPE on the test set, the latter model fits the data better with a $16\%$ error, instead of the $18\%$ for the first model.
%
%Parameters roots for both models have values within the unit circle.

@@

\subsection{Lockdown influence}
Even if the two models have the same order, they cannot be directly compared using the Information Criteria, for the Membach model works after a transformation pass, and they are fitted from different data.

@@

As we can see in \figurename~\ref{fig:forecast.lockdown}, at both stations the models overestimate starting after the lockdown. Uccle data in particular show a greater overestimation, that can also be noticed by looking at the difference between the forecasted and measured values.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/final-model-5"}
		\caption{Uccle station}
		\label{fig:forecast.lockdown:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/final-model-5"}
		\caption{Membach station}
		\label{fig:forecast.lockdown:mems}
	\end{subfigure}
	\caption{Forecasts from March $2^{nd}$.}{\centering\small The forecasted values are on the background. March $14^{th}$ at midnight is highlighted.\\}
	\label{fig:forecast.lockdown}
\end{figure}
%
We assess analytical values obtained from the new 8 weeks forecast:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item the MAPE for Uccle triples with respect to the original test set to $\approx46\%$, and only 39\% and 63\% of the test data falls inside the confidence intevals (80\% and 95\%).
	\item regarding the Membach station forecasts, the MAPE is reduced to $14\%$, while 82\% and 96\% of data falls inside the confidence intervals.
\end{itemize}

@@

Checking the seasonal plots of the complete weekly seasonality (\figurename~\ref{fig:season.lockdown:ms:mems}), it is possible to notice a flattening of the values towards 0, in particular regarding Monday to Friday values.

@@

A second mean to address the multi-seasonalities is through a harmonic regression approach. Using that we could model multiple seasonalities at once through Fourier terms. One difficulty we could encounter is that our weekly seasonality appear to be flat for most of the week (\figurename~\ref{fig:season.lockdown}), and such a behaviour is not easy to model with a few Fourier terms per seasonality, leading to computationally heavy modelling and forecasting.

Different models, such as STL (Seasonal and Trend decomposition using Loess) or TBATS (Trigonometric seasonality, Box-Cox transformation, ARMA errors, Trend and Seasonal components), are suggested in the literature \cite[chapter 12]{hyndman2018forecasting} to address multiple and complex seasonalities, also when they change in time. These could also better tackle the missing Normality assumption of our data.

>>>>

The correct estimation of a background noise, correlated or not with human activity, could help institutions to better isolate less prominent earthquakes and similar events from the seismic data. This could result in better understanding of the frequency of earthquakes by studying their more precise evolution over time.

\subsubsection{Data source and description}
The Royal Observatory of Belgium (ROB) gathers and publishes \cite{RoyalDataPolicy,Data20200402} data about vertical ground displacement from 9 stations through the Country, linking them to seismic events whenever possible \cite{DataEvents20200402}. They also offer a data visualization tool for the same data \cite{DataVisualization20200402}.

@@

Out of the 9 stations, the following 6 have interesting positions, related to possible correlations with human activity:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item \textbf{Uccle}, \textbf{Sart-Tilman}, and \textbf{Ében-Émael} are near a city (Brussels, Liege, and Maastricht).
	\item \textbf{Ostenda} is near the city of Ostenda, that borders with the sea.
	\item \textbf{Membach} and \textbf{Dourbes} are in towns near a natural park.
\end{itemize}

@@

\subsection{Data cleanup and transformation}\label{sec:cleanup-transform}
\subsubsection{Data gathering and cleanup}
The data is provided in CSV format, one file per day, with only two columns related to the min-max pair \cite{Data20200402} and no column related to time. This causes two problems:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item For merging days, we need to add a column with the second-precise datetime
	\item In days with missing values there are placeholder values instead of a missing row, that would make the following rows misinterpreted for measurements at the wrong second. The placeholder values are one of these two min-max pairs: $(0,0),\, (-1,1)$.
\end{enumerate}

@@

\subsubsection{Data aggregation and initial transformation}
Since we are not interested in a second-precise analysis, and it would be computationally expensive to deal with the whole dataset, we aggregate the data by hour. The aggregation results in around 14 thousand points, that can still be captured by interpretable models.

While performing this operation, we should consider the correct time-zone: as we anticipated, any seasonal analysis dealing with UTC dates would be out of phase for the half-year of legal time, with respect to the half-year of solar time. Employing the \texttt{lubridate} R package we converted times to local time-zones, and then we kept them as absolute instants, because all the R functions we use work on the UTC conversion of datetimes by default. This transformation made it possible to have each hour in the timeseries to represent the local time in Belgium (\textit{Europe/Brussels}), without being influenced by the legal time.

The aggregation was performed considering the mean total ground displacement at each hour, for the following reasons:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item When an hour contains some missing seconds, these do not heavily affect the aggregation result, in contrast with summing the values
	\item Seismic events, such as quakes and mine explosions, normally last some minutes at most, and will affect the result only slightly, as opposed to the hourly maximum
	\item Usually, maximum value aggregation does not follow the Normal distribution, while some models that we are going to use assume the data follow a Normal distribution
\end{itemize}

@@

Since one of the main focuses is on the seasonalities themselves, we aim to avoid searching for correlation with human activity while imposing human-based seasonality periods. Thus, we use an analytical method to extract seasonalities that the data itself suggests to us. Still, we recognize that the already performed transformations impose an human-based time alteration to the data itself (Section \ref{sec:cleanup-transform}).

The method we are going to apply is based on the Fourier Transform (FT), and generates a \textit{periodgram} highlighting the frequencies that might correspond to a seasonality component \cite[chapter 11]{brockwell1991time}.
Applying the FT for any frequency, we transform our 1-dimensional data into 2-dimensional data mapped around a circle. Then, we can sum all the resulting vectors (that are our new data points) to obtain a vector for each frequency, whose absolute value represents the power level of the same frequency. An high power level represents a significant periodicity on that frequency in our original data, since it means that the corresponding 2-dimensional data behaves in a highly similar way during each cycle.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/whole data plot-2"}
		\caption{Uccle station}
		\label{fig:zoom:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/whole data plot-2"}
		\caption{Membach station}
		\label{fig:zoom:mems}
	\end{subfigure}
	\caption{Zoom from January $6^{th}$ to February $3^{rd}$ 2019}
	\label{fig:zoom}
\end{figure}
%
A common operation to increase the difference between significant and non-significant periodicities is to apply a Moving Average (MA) smoothing filter with a smaller window than the interesting ones. We will use such a filter with $p=2$.

On this note we must specify that, when a significant periodicity is found, it influences smaller periodicities given by $f*\{2,3,\dots\}$ the significant frequency \cite[chapter 11]{brockwell1991time}. Also, if any periodicity is found that is longer than the original dataset, it must be considered as simple chance, rather than serendipity.

@@

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/periodgram-1"}
		\caption{Uccle station}
		\label{fig:periodgram:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/periodgram-1"}
		\caption{Membach station}
		\label{fig:periodgram:mems}
	\end{subfigure}
	\caption{Periodgram plots}
	\label{fig:periodgram}
\end{figure}
%
Both plots highlight multiple 24-hour periodicities. This is given by approximation factors, since we round the divisions result to the nearest natural number. It also highlights that using a single 24-hour seasonality could leave some information on the residuals. Still, in both stations it is the most significant periodicity.

@@

\subsubsection{Dataset division}
As best practices suggest, we divide the dataset into training and testing datasets with a ratio around $85\%/15\% - 75\%/25\%$.
Since one of our objectives is to assess any significant difference from before and after the Belgium lockdown (on March $14^{th}$ the \textit{soft} lockdown, while on March $18^{th}$ the complete lockdown), we selected data from the start of our timeseries until 20-01-2020 (64 weeks $\approx 82\%$) as training, while the test follows ending on 27-04-2020 (14 weeks $\approx 18\%$).

@@

This initial division demostrated to be computationally heavy while fitting some models with weekly seasonality, so that it would be challenging to fit more than a couple of models.
Given this issue, we choose the training/testing sets to hold 52/18 weeks: this is the greater train-to-test ratio we deem usable ($\approx 75\%/25\%$), while maintaining computational feasibility.
The assessment of the lockdown influence will be discussed in Section~\ref{sec:methodology:lockdown}.

Some R functions cannot handle timeseries with missing values, but as we have seen both stations' datasets include missing values.
Also, please note that we are going to have a timeseries object indexed by weeks, so that it contains the 168 hours between each integer index.
Since we know that the longest missing values series is 90 hours long, we can apply a simple completion strategy for serving those commands a filled timeseries: we will substitute missing values with the mean of the week they are in; in R, \texttt{na.aggregate(x, floor)}.

\subsubsection{Decomposition}
In order to decompose our dataset into trend and seasonality(-ties) components, we followed these steps:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply a smoothing filter to extract the trend component
	\item Remove the trend from the dataset and extract the seasonal figure using the local mean method
	\item Remove the trend and the seasonal components to obtain the residuals
\end{enumerate}
While doing so, we tested different smoothing filters: simple filter (p=seasonality), Spencer's 15-point filter and MA smoothing filter (p=seasonality). Since they do not produce particularly different seasonalities, we choose to keep the results obtained by applying the MA smoothing filter, for its significance in terms of seasonally adjusted data.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/extract both 24 and 168-1.pdf}
		\caption{Uccle station}
		\label{fig:decomposed:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.mems_files/figure-latex/extract both 24 and 168-1.pdf}
		\caption{Membach station}
		\label{fig:decomposed:mems}
	\end{subfigure}
	\caption{Timeseries decomposition}{\centering\small Only the second method results are shown for clarity. Week 10 includes January $1^{st}$.\\\centering x-axes on left columns represent weeks, on right columns represent hours.\\}
	\label{fig:decomposed}
\end{figure}
%
Since the periodgrams suggested the presence of two significative periodicities, we also tested the extraction of both corresponding seasonalities (\figurename~\ref{fig:decomposed}):
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply the MA smoothing filter (p=24) on the dataset to extract the trend component
	\item Remove the trend from the dataset and extract the 24-hour seasonal figure using the local mean method
	\item Remove the 24-hour seasonal components to obtain a deseasoned dataset
	\item Apply the MA smoothing filter (p=168) on the deseasoned dataset to extract the updated trend component
	\item Remove the trend from the already deseasoned dataset and extract the 168-hour seasonal figure using the local mean method
	\item Repeat steps 1-5 using the last deseasoned dataset until convergence (2-3 cycles)
	\item Removing the last fitted trend and the two seasonalities we obtain the final residuals
\end{enumerate}

@@

From the same decomposition results, we notice that data around January $1^{st}$ show a significant drop in values. That is highlighted by both trends, where the overall minimum falls in that period, and by the residuals for Uccle data, showing a clear overestimation.

\section{Methodology}\label{sec:methodology}
Having two different datasets, we will operate on them separately and then we will compare the model interpretations.

@@

The model we fit is the ARIMA\footnote{AutoRegressive Integrated Moving Average} model, including the seasonality components. To identify the order of the different components described by ARIMA, we will analyse the autocorrelation (ACF) and partial autocorrelation (PACF) functions of our datasets.
As a reminder:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item non-seasonal AR process of order $p$ is described by an exponentially or sigmoidally decaying ACF, and by a cut-off of significant PACF values at lag $p$
	\item non-seasonal MA process of order $q$ is described by an exponentially or sigmoidally decaying PACF, and by a cut-off of significant ACF values at lag $q$
	\item seasonal AR process of order $P$ is described by an exponential or sigmoidal decay at seasonal lags on ACF values, and by a cut-off of significant PACF values at each seasonal lag until the $P^{th}$
	\item seasonal MA process of order $Q$ is described by an exponential or sigmoidal decay at seasonal lags on PACF values, and by a cut-off of significant ACF values at each seasonal lag until the $Q^{th}$
	\item given how significance is assessed for the ACF/PACF spikes, one over 20 spikes (5\%) could be \textit{coincidentally significant} or \textit{coincidentally \textbf{non}-significant}
\end{itemize}

The same model will be then used to forecast other 8 weeks from March $2^{nd}$, so that we can assess the forecast power over the lockdown period. Given the fact that both our series are quite stable over time (\figurename{s} \ref{fig:decomposed:uccs}, \ref{fig:mems:box-cox:data}), we can expect that the predicted variability for further forecasts will not significantly increase, compromising the long-term forecasts.

@@

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-35}
		\caption{ACF and PACF plots}
		\label{fig:arima-26}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-32}
		\caption{Q-Q plot}
		\label{fig:arima-24}
	\end{subfigure}
	\caption{Plots for $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
	\label{fig:arima}
\end{figure}
%
The ACF and PACF of the residuals for all three models are very similar (\figurename~\ref{fig:arima-26}), even if the models themselves have great differences in meaning and in number of parameters: they highlight a clear seasonal ARMA process with periodicity 24. This shows how our previous conclusion that the 168 hours seasonality could include the 24 hours one is flawed. Unfortunately, how we are using the R function do not permit us to estimate multiple seasonalities\footnote{\textit{Arima} method from the \textit{forecast} R package, only specifying the seasonal and non-seasonal orders.}.

@@

Also, it must be noted that a January $1^{st}$ period lies in our test dataset, that was already noticed as low-valued period (\figurename~\ref{fig:decomposed}). We can notice a clear overestimation during those couple of weeks (around week 62 on \figurename~\ref{fig:arima-forecast}).

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{project.uccs_files/figure-latex/arima-37}
		\caption{$\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
		\label{fig:arima-forecast:200011}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{project.uccs_files/figure-latex/arima-69}
		\caption{$\mathrm{ARIMA}(3,1,1)(0,1,1)_{168}$}
		\label{fig:arima-forecast:412011}
	\end{subfigure}
	\caption{ARIMA forecasts}{\centering\footnotesize$\mathrm{ARIMA}(4,1,2)(0,1,1)_{168}$ behaves like $\mathrm{ARIMA}(3,1,1)(0,1,1)_{168}$ with faster increasing variablity\\}
	\label{fig:arima-forecast}
\end{figure}
%
We can see in \figurename~\ref{fig:arima-forecast} the forecast behaviour for our models. We can appreciate the constant span of the confidence intervals in \figurename~\ref{fig:arima-forecast:200011}, in constrast with the more complex models that have increasing intervals further away from the training set. It must be noted that the increasing intervals include negative values, that are actually impossible to obtain given the data meaning (difference between maximum and minimum measurements).

@@

\subsection{Membach station}
\subsubsection{ARIMA}
Similarly to data from Uccle, data from Membach does not respect the Normality assumption. We tested if the Q-Q plot showed significantly better approximation of the Normal distribution after applying the \textit{log} or \textit{Box-Cox} transformations, and the latter makes the data approach Normality with $\lambda\approx-1$ (\figurename~\ref{fig:mems:box-cox}). This solves the high peaks problem we already noticed while decomposing to highlight the trend and seasonalities.

@@

The employment of \textit{auto.arima} stepwise methodology$^{\ref{footnote:auto.arima}}$ highlights the first model improved with one more non-seasonal MA parameter. This is confirmed by the ACF/PACF plots of its residuals, while their analysis for the second model do not lead to any improvement.
The ACF and PACF plots of both models are similar to the ones from the chosen model for the Uccle station (\figurename~\ref{fig:arima-24}), and we can notice the same 24 hours remaining periodicity.

@@

Analytically, we have quite a difference between the two models in terms of adherence of the confidence intervals:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item $\mathrm{ARIMA}(0,1,1)(0,1,1)_{168}$: $\approx98\%$ of data is in the $80\%$ CI and $\approx99\%$ in the $95\%$ CI
	\item $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$: $\approx73\%$ of data is in the $80\%$ CI and $\approx89\%$ in the $95\%$ CI
\end{itemize}
As we see, the first model over-generalises the variability, while the latter seems more precise, even if losing some information. Considering the MAPE on the test set, the latter model fits the data better with a $16\%$ error, instead of the $18\%$ for the first model.
%
%Parameters roots for both models have values within the unit circle.

@@

\subsection{Lockdown influence}\label{sec:methodology:lockdown}
Even if the two models have the same order, they cannot be directly compared using the Information Criteria, for the Membach model works after a transformation pass, and they are fitted from different data.

@@

As we can see in \figurename~\ref{fig:forecast.lockdown}, at both stations the models overestimate starting after the lockdown. Uccle data in particular show a greater overestimation.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/final-model-5"}
		\caption{Uccle station}
		\label{fig:forecast.lockdown:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/final-model-5"}
		\caption{Membach station}
		\label{fig:forecast.lockdown:mems}
	\end{subfigure}
	\caption{Forecasts from March $2^{nd}$.}{\centering\small The forecasted values are on the background. March $14^{th}$ at midnight is highlighted.\\}
	\label{fig:forecast.lockdown}
\end{figure}
%
We assess analytical values obtained from the new 8 weeks forecast:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item the MAPE for Uccle triples with respect to the original test set to $\approx46\%$, and only 39\% and 63\% of the lockdown data falls inside the confidence intevals (80\% and 95\%).
	\item regarding the Membach station forecasts, the MAPE is reduced to $14\%$, while 82\% and 96\% of data falls inside the confidence intervals.
\end{itemize}

@@

Checking the seasonal plots of the complete weekly seasonality (\figurename~\ref{fig:season.lockdown:ms:mems}), it is possible to notice a flattening of the values towards 0, in particular regarding Monday to Friday values. This seem to confirm the initial hypotesis of lockdown influence.

@@

A second mean to address the multi-seasonalities is through a harmonic regression approach. Using that, we could model multiple seasonalities at once through Fourier terms. One difficulty we could encounter is that our weekly seasonality appear to be flat for most of the week (\figurename~\ref{fig:season.lockdown}), and such a behaviour is not easy to model with a few Fourier terms per seasonality, leading to computationally heavy modelling and forecasting.

Different models, such as STL (Seasonal and Trend decomposition using Loess) or TBATS (Trigonometric seasonality, Box-Cox transformation, ARMA errors, Trend and Seasonal components), are suggested in the literature \cite[chapter 12]{hyndman2018forecasting} to address multiple and complex seasonalities, also when they change in time. These could better tackle the missing Normality assumption of our data, along with relaxing the initial transformation between timezones.
